---
title: "Injection Data Analysis Code for Public Use"
author: "Emma"
date: "2025-12-15"
output: html_document
---

```{r library loading}
library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
library(mgcv)
library(ggplot2)
library(lmtest)
library(lme4)
library(betareg)
library(survival)
library(survminer)
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load and clean data}
injection <- readr::read_csv(here("data", "inputs", "injection_trial_data.csv"))

injection <- injection %>%
  rename(
    Assay_Day = `Assay Day`,
    CPB_Group = `CPB Group`,
    Defoliation = `% Defoliation`
  )

# average beetle weight & sd in each rep 
injection <- injection %>%
  mutate(
    GroupPrefix = str_extract(CPB_Group, "^[^\\.]+"),
    `Weight (g)` = as.numeric(`Weight (g)`),
    Defoliation = as.numeric(Defoliation),
    avg_beetle_weight = ifelse(Alive > 0, `Weight (g)` / Alive, NA),
    defol_per_beetle = ifelse(Alive > 0, Defoliation / Alive, NA)
  )

# summarize by GroupPrefix
injection_group_summary <- injection %>%
  group_by(GroupPrefix) %>%
  summarize(
    n_reps = n(),
    mean_avg_beetle_weight = mean(avg_beetle_weight, na.rm = TRUE),
    sd_avg_beetle_weight = sd(avg_beetle_weight, na.rm = TRUE),
    mean_defol_per_beetle = mean(defol_per_beetle, na.rm = TRUE),
    sd_defol_per_beetle = sd(defol_per_beetle, na.rm = TRUE),
    .groups = "drop"
  )

# join group summary back into dish-level data
injection <- injection %>%
  left_join(injection_group_summary, by = "GroupPrefix", suffix = c("", "_GroupAvg"))

# extract prefix and assign treatments for Injection Trial
injection <- injection %>%
  mutate(GroupPrefix = str_extract(CPB_Group, "^[^\\.]+")) %>%
  mutate(Treatment = case_when(
    GroupPrefix == "C1" ~ "Control_Uninjected",
    GroupPrefix == "C2" ~ "Control_H2O",
    GroupPrefix == "C3" ~ "Control_dsRNA",
    GroupPrefix == "GD1" ~ "GPA2_Low",
    GroupPrefix == "GD2" ~ "GPA2_High",
    GroupPrefix == "LD1" ~ "LGR1_Low",
    GroupPrefix == "LD2" ~ "LGR1_High",
    TRUE ~ "Other"
  ))

table(injection$Treatment)

injection <- injection %>%
  mutate(Treatment = relevel(factor(Treatment), ref = "Control_Uninjected"))

```

# Weight Analyses:

```{r compare model fits for weight}
# let's compare four fits - linear, quadratic, quartic, and means at different points

# model 1: linear------------------------------------------------------
lmm_injection <- lmer(avg_beetle_weight ~ Treatment * Assay_Day + (1 | CPB_Group), data = injection)
summary(lmm_injection)
anova(lmm_injection)

par(mfrow = c(2,2))
plot(lmm_injection)

# linear visualization
injection %>% 
  ggplot(aes(x = Assay_Day, y = avg_beetle_weight, color = Treatment, group = CPB_Group)) + 
  geom_point() +
  xlim(c(0,8)) + 
  # geom_line()
  # geom_smooth(se = F)
  geom_smooth(method = "lm", se = F)

injection %>% 
  ggplot(aes(x = Assay_Day, y = avg_beetle_weight, color = Treatment, group = CPB_Group)) + 
  geom_point() +
  xlim(c(0,8))

# loess visualization
injection %>% 
  ggplot(aes(x = Assay_Day, y = avg_beetle_weight, color = Treatment, group = CPB_Group)) + 
  geom_point() +
  xlim(c(0,8)) + 
  # geom_line()
  # geom_smooth(se = F)
  geom_smooth( se = F, lwd = 0.25)

# model 2: quadratic------------------------------------------------------
lmm_quad_injection <- lmer(avg_beetle_weight ~ Treatment *  (Assay_Day + I(Assay_Day^2)) + (1 | CPB_Group) + (1 | CPB_Group), data = injection)
summary(lmm_quad_injection)
anova(lmm_quad_injection)

plot(lmm_quad_injection)

# model 3: quartic------------------------------------------------------
lmm_quartic_injection <- lmer(avg_beetle_weight ~ Treatment *   (Assay_Day + I(Assay_Day^2) + I(Assay_Day^3) + I(Assay_Day^4)) + (1 | CPB_Group), data = injection)
summary(lmm_quartic_injection)
anova(lmm_quartic_injection)

plot(lmm_quartic_injection)

# model 4: time as factor------------------------------------------------------
lmm_time_factor_injection <- lmer(avg_beetle_weight ~ Treatment *  as.factor(Assay_Day) + (1 | CPB_Group), data = injection)
summary(lmm_time_factor_injection)
anova(lmm_time_factor_injection)

plot(lmm_time_factor_injection)

# time as factor visualization
injection %>% 
  ggplot(aes(x = Assay_Day, y = avg_beetle_weight, color = Treatment, group = CPB_Group)) + 
  geom_point() + 
  geom_line() + 
  xlim(c(0,8))

# model 5: non-linear------------------------------------------------------
nonlinear_ls_model <-
  nls(avg_beetle_weight ~ 1 - exp(-alpha*((Assay_Day)^2)),
      data = injection,
      start = list(alpha = 1))
summary(nonlinear_ls_model)

plot(nonlinear_ls_model)

# it appears that a polynomial model is a poor fit, and a linear model is too simple

# compare the BICs, the lowest BIC is best
paste0("Linear model BIC is ", BIC(lmm_injection))
paste0("Quadratic model BIC is ", BIC(lmm_quad_injection))
paste0("Quartic model BIC is ", BIC(lmm_quartic_injection))
paste0("Time as factor model BIC is ", BIC(lmm_time_factor_injection))
paste0("Linear model BIC is ", BIC(nonlinear_ls_model))

# model 6: GAM------------------------------------------------------
gam_injection_model <-  gam(avg_beetle_weight ~ CPB_Group+s(Assay_Day, k = 3), data = injection)

BIC(gam_injection_model)

# add predictions to original dataset
injection <- injection %>%
  mutate(prediction_gam = predict(gam_injection_model, newdata = injection))

# plot observed and predicted values
ggplot(injection, aes(x = Assay_Day, y = avg_beetle_weight, color = Treatment, group = CPB_Group)) +
  geom_point() +
  geom_line(aes(y = prediction_gam)) +
  xlim(0, 8) +
  labs(title = "GAM Fit: Avg Beetle Weight over Time", y = "Avg Beetle Weight (g)") +
  theme_minimal()

# This seems to fit the data well. There is probably a way to make this slightly better
# by adjusting the number of knots (k), or considering different basis functions.
# Cubic splines are slightly more interpretable. 

# We need to sort the data in order of CBP Group, then Date.
# This will not change the model, but will enable us to do a Durbin-Watson test

injection_gam_sort <- injection %>% group_by(CPB_Group, Assay_Day) %>% 
  arrange(CPB_Group, Assay_Day) %>% 
  filter(!is.na(avg_beetle_weight))

gam_injection_model_2 <-  gam(avg_beetle_weight ~ CPB_Group+s(Assay_Day, k = 3), data = injection_gam_sort)

dwtest(gam_injection_model_2$residuals ~ 1)

# DW ~ 2.00, so no autocorrelation in residuals and non-signficant p-value so we fail to reject the null hypothesis of no autocorrelation
```

```{r test for significance of variables in weight}
# now we move on to testing for significant differences in weight over time between treatment groups

# model 1: full parametruc using REML
lmm_reml_inj <- lmer(
  avg_beetle_weight ~ Treatment * Assay_Day + (1 | CPB_Group),
  data = injection,
  REML = TRUE  #default: gives unbiased estimates of random effects
)

# model 1: full parametruc using REML
lmm_ml_inj <- lmer(
  avg_beetle_weight ~ Treatment * Assay_Day + (1 | CPB_Group),
  data = injection,
  REML = FALSE  #default: gives unbiased estimates of random effects
)

summary(lmm_reml_inj)
summary(lmm_ml_inj)

# refit the model using ML (REML = FALSE) becuase it is required for comparing models using anova
# lmm_ml_inj <- update(lmm_reml, REML = FALSE)

# model 2: nested/reduced model with ML
    #remove fixed effects or interaction terms to test their significance
lmm_reduced_ml_inj <- lmer(
  avg_beetle_weight ~ Treatment + Assay_Day + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

# compare models using Likelihood Ratio Test (LRT)
anova(lmm_reduced_ml_inj, lmm_ml_inj)

# model 3: nested/reduced model with ML
    #remove fixed effects or interaction terms to test their significance
lmm_reduced_ml_noday <- lmer(
  avg_beetle_weight ~ Treatment + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

anova(lmm_reduced_ml_noday, lmm_ml_inj)

# model 4: nested/reduced model with ML
    #remove fixed effects or interaction terms to test their significance
lmm_reduced_ml_notreatment <- lmer(
  avg_beetle_weight ~ Assay_Day + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

anova(lmm_reduced_ml_notreatment, lmm_ml_inj)

anova(lmm_reduced_ml_notreatment, lmm_reduced_ml_inj)

summary(lmm_reml_inj)
```

```{r visualizations for weight}
injection_palette <- c(
  "Control_Uninjected" = "#A8E6A3",
  "Control_H2O"        = "#81C784",
  "Control_dsRNA"      = "#4CAF50",
  "GPA2_Low"           = "#81D4FA",
  "GPA2_High"          = "#0288D1",
  "LGR1_Low"           = "#CE93D8",
  "LGR1_High"          = "#8E24AA"
)

injection_gam_sort <- injection_gam_sort %>%
  select(-Treatment) %>%  # remove broken NA column, just in case
  left_join(
    injection %>% select(CPB_Group, Treatment) %>% distinct(),
    by = "CPB_Group"
  )

injection_gam_sort$Treatment <- factor(injection_gam_sort$Treatment, levels = names(injection_palette))

# summary for treatment-level mean and SE
inj_gam_weight_summary <- injection_gam_sort %>%
  group_by(Treatment, Assay_Day) %>%
  summarise(
    mean_pred = mean(prediction_gam, na.rm = TRUE),
    se_pred = sd(prediction_gam, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

inj_gam_weight_summary$Treatment <- factor(inj_gam_weight_summary$Treatment, levels = names(injection_palette))
injection_gam_sort$Treatment <- factor(injection_gam_sort$Treatment, levels = names(injection_palette))

ggplot(inj_gam_weight_summary, aes(x = Assay_Day, y = mean_pred)) +
  geom_ribbon(aes(fill = Treatment, ymin = mean_pred - se_pred, ymax = mean_pred + se_pred),
              alpha = 0.2, color = NA) +
  geom_line(aes(color = Treatment), size = 1.2) +
  scale_color_manual(values = injection_palette) +
  scale_fill_manual(values = injection_palette) +
  labs(
    x = "Assay Day",
    y = "Predicted Avg Beetle Weight (g)",
    title = "Weight Over Time by Treatment"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

# plot individaul lines per treatment
injection$Treatment <- factor(injection$Treatment, levels = names(injection_palette))
inj_gam_weight_summary$Treatment <- factor(inj_gam_weight_summary$Treatment, levels = names(injection_palette))

ggplot() +
  geom_line(
    data = injection %>%
      filter(Assay_Day <= max(inj_gam_weight_summary$Assay_Day)),
    aes(x = Assay_Day, y = prediction_gam, group = CPB_Group, color = Treatment),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_ribbon(
    data = inj_gam_weight_summary,
    aes(x = Assay_Day, ymin = mean_pred - se_pred, ymax = mean_pred + se_pred, fill = Treatment),
    alpha = 0.2,
    color = NA
  ) +
  geom_line(
    data = inj_gam_weight_summary,
    aes(x = Assay_Day, y = mean_pred, color = Treatment),
    size = 1.2
  ) +
  scale_color_manual(
    values = injection_palette,
    aesthetics = c("color", "fill")
  ) +
  labs(
    x = "Assay Day",
    y = "Average Beetle Weight (g)",
    title = "Weight Over Time by Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

# Defoliation Analyses:

```{r compare model fits for defoliation}
# center time variable
injection <- injection %>%
  mutate(Assay_Day_c = scale(Assay_Day, center = TRUE, scale = FALSE))

# linear
lmm_defol_linear_inj <- lmer(Defoliation ~ Treatment * Assay_Day_c + Alive + (1 | CPB_Group), data = injection)

# quadratic
lmm_defol_quad_inj <- lmer(Defoliation ~ Treatment * (Assay_Day_c + I(Assay_Day_c^2)) + Alive + (1 | CPB_Group), data = injection)

# quartic
lmm_defol_quartic_inj <- lmer(Defoliation ~ Treatment * (Assay_Day_c + I(Assay_Day_c^2) + I(Assay_Day_c^3) + I(Assay_Day_c^4)) + Alive + (1 | CPB_Group), data = injection)

# time as factor
lmm_defol_time_factor_inj <- lmer(Defoliation ~ Treatment * as.factor(Assay_Day) + Alive + (1 | CPB_Group), data = injection)

# GAM
injection$CPB_Group <- as.factor(injection$CPB_Group)

gam_defol_inj <- gam(
  Defoliation ~ Treatment + s(Assay_Day, by = Treatment, k = 7) + Alive + s(CPB_Group, bs = "re"),
  data = injection
)

gam_defol2_inj <- gam(
  Defoliation ~ Treatment + s(Assay_Day, by = Treatment, k = 4) + Alive + s(CPB_Group, bs = "re"),
  data = injection
)

# non-linear least squares
nls_defol_inj <- nls(
  Defoliation ~ Asym * (1 - exp(-k * Assay_Day)),
  data = injection,
  start = list(Asym = 80, k = 1),
  control = nls.control(minFactor = 1e-10, warnOnly = TRUE)
)

# beta regression
injection <- injection %>%
  mutate(defol_01 = (Defoliation + 0.001) / 100.002) #standard fix for this regression, Smithson & Verkuilen adjustment I think?

beta_defol_inj <- betareg(defol_01 ~ Treatment * Assay_Day + Alive, data = injection)
summary(beta_defol_inj)

# compare BICs across the models
cat("Linear BIC:", BIC(lmm_defol_linear_inj), "\n") #1868.854
cat("Quadratic BIC:", BIC(lmm_defol_quad_inj), "\n") #1846.146
cat("Quartic BIC:", BIC(lmm_defol_quartic_inj), "\n") #1822.12
cat("Time-as-factor BIC:", BIC(lmm_defol_time_factor_inj), "\n") #1655.63
cat("GAM BIC:", BIC(gam_defol_inj), "\n") #1787.931
cat("GAM2 BIC:", BIC(gam_defol2_inj), "\n") #1787.357
cat("NLS BIC:", BIC(nls_defol_inj), "\n") #2005.965
cat("Beta BIC:", BIC(beta_defol_inj), "\n") #-973.7633

# Time-as-factor is the best parametric model but Beta-Regression performed best by absolute BIC. Testing for signifcance with Time-as-factor and will vosualize both to see which represents the data better.

# compare simple and complex models
# full interaction model
lmm_defol_tf_full <- lmer(
  Defoliation ~ Treatment * as.factor(Assay_Day) + Alive + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

# additive model (no interaction)
lmm_defol_tf_add <- lmer(
  Defoliation ~ Treatment + as.factor(Assay_Day) + Alive + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

# treatment-only model
lmm_defol_tf_treat <- lmer(
  Defoliation ~ Treatment + Alive + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

# day-only model
lmm_defol_tf_day <- lmer(
  Defoliation ~ as.factor(Assay_Day) + Alive + (1 | CPB_Group),
  data = injection,
  REML = FALSE
)

# likelihood ratio tests
# test for interaction 
anova(lmm_defol_tf_add, lmm_defol_tf_full)
    #defoliation patterns vary over time depending on treatment

# test for day effect
anova(lmm_defol_tf_treat, lmm_defol_tf_full)
    #defoliation changes significantly over days, no matter the treatment

# test for treatment effect
anova(lmm_defol_tf_day, lmm_defol_tf_full)
    #defoliation differs between treatments

# test for treatment effect in additive model
anova(lmm_defol_tf_day, lmm_defol_tf_add)
    #treatment not significant once day is included
    #treatment differences emerge bexause of different time trajectories, not as constant offsets

##use REML = TRUE for reporting
```

```{r visaulizations for defoliation}
# defoliation: time-as-factor
inj_timeasfactor_data <- model.frame(lmm_defol_time_factor_inj)

inj_timeasfactor_data$predicted_defol <- predict(lmm_defol_time_factor_inj)

inj_timeasfactor_data <- inj_timeasfactor_data %>%
  rename(Assay_Day = `as.factor(Assay_Day)`) %>%
  mutate(Assay_Day = as.numeric(as.character(Assay_Day)))

inj_timeasfactor_summary <- inj_timeasfactor_data %>%
  group_by(Treatment, Assay_Day) %>%
  summarise(
    mean_pred = mean(predicted_defol, na.rm = TRUE),
    se_pred = sd(predicted_defol, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

inj_timeasfactor_data$Treatment <- factor(inj_timeasfactor_data$Treatment,
  levels = names(injection_palette)
)

inj_timeasfactor_summary$Treatment <- factor(inj_timeasfactor_summary$Treatment,
  levels = names(injection_palette)
)

ggplot(inj_timeasfactor_summary, aes(x = Assay_Day, y = mean_pred, color = Treatment, fill = Treatment)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = mean_pred - se_pred, ymax = mean_pred + se_pred), alpha = 0.2, color = NA) +
  scale_color_manual(values = injection_palette) +
  scale_fill_manual(values = injection_palette) +
  labs(
    x = "Assay Day",
    y = "Model-Predicted Defoliation (%)",
    title = "Defoliation Over Time by Treatment"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

# plot: model predictions with raw data points
ggplot(inj_timeasfactor_summary, aes(x = Assay_Day, y = mean_pred, color = Treatment, fill = Treatment)) +
  # raw data points
  geom_jitter(
    data = inj_timeasfactor_data,
    aes(x = Assay_Day, y = Defoliation, color = Treatment),
    alpha = 0.3,
    width = 0.2,
    size = 1,
    inherit.aes = FALSE
  ) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = mean_pred - se_pred, ymax = mean_pred + se_pred),
              alpha = 0.2, color = NA) +
  scale_color_manual(values = injection_palette) +
  scale_fill_manual(values = injection_palette) +
  labs(
    x = "Assay Day",
    y = "Defoliation (%)",
    title = "Defoliation Over Time by Treatment"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

# individual dish lines per treatment
ggplot() +
  geom_line(
    data = inj_timeasfactor_data,
    aes(x = Assay_Day, y = predicted_defol, group = CPB_Group, color = Treatment),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_ribbon(
    data = inj_timeasfactor_summary,
    aes(x = Assay_Day, ymin = mean_pred - se_pred, ymax = mean_pred + se_pred, fill = Treatment),
    alpha = 0.2,
    color = NA
  ) +
  geom_line(
    data = inj_timeasfactor_summary,
    aes(x = Assay_Day, y = mean_pred, color = Treatment),
    size = 1.2
  ) +
  
  # palette and labels
  scale_color_manual(values = injection_palette) +
  scale_fill_manual(values = injection_palette) +
  labs(
    x = "Assay Day",
    y = "Defoliation (%)",
    title = "Defoliation Over Time by Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )
```
```{r beta regression for defoliation and visualization}
library(dplyr)
library(betareg)

# only using rows that actually have defoliation recorded
inj_defol <- injection %>%
  filter(!is.na(Defoliation), !is.na(Assay_Day), !is.na(Treatment), !is.na(Alive)) %>%
  mutate(
    defol_01 = (Defoliation + 0.001) / 100.002,     
    Treatment = factor(Treatment, levels = names(injection_palette))
  )

beta_defol_inj <- betareg(defol_01 ~ Treatment * Assay_Day + Alive, data = inj_defol)
summary(beta_defol_inj)

# predictions on the same rows (response scale 0–1)
inj_defol <- inj_defol %>%
  mutate(pred_beta = predict(beta_defol_inj, newdata = inj_defol, type = "response"))

# build summaries in % units
#mean from model preds; SE from observed proportions
inj_beta_summary <- inj_defol %>%
  group_by(Treatment, Assay_Day) %>%
  summarise(
    mean_pred = mean(pred_beta, na.rm = TRUE) * 100,                       # %
    se_pred   = (sd(defol_01, na.rm = TRUE) / sqrt(n())) * 100,            # %
    .groups   = "drop"
  )

inj_timeasfactor_data <- inj_defol %>%
  mutate(predicted_defol = pred_beta * 100)  # %

inj_timeasfactor_summary <- inj_beta_summary

#individual dish lines per treatment
ggplot() +
  geom_line(
    data = inj_timeasfactor_data,
    aes(x = Assay_Day, y = predicted_defol, group = CPB_Group, color = Treatment),
    alpha = 0.3,
    size = 0.6
  ) +
  geom_ribbon(
    data = inj_timeasfactor_summary,
    aes(x = Assay_Day, ymin = mean_pred - se_pred, ymax = mean_pred + se_pred, fill = Treatment),
    alpha = 0.2,
    color = NA
  ) +
  geom_line(
    data = inj_timeasfactor_summary,
    aes(x = Assay_Day, y = mean_pred, color = Treatment),
    size = 1.2
  ) +
  scale_color_manual(values = injection_palette) +
  scale_fill_manual(values = injection_palette) +
  labs(
    x = "Assay Day",
    y = "Defoliation (%)",
    title = "Defoliation Over Time by Treatment (Beta Regression)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "right")

```

# Survival Analyses:

```{r kaplan-meier survival analysis}
# expand rows for dead beetles
inj_death_rows <- injection %>%
  filter(Dead > 0) %>%
  uncount(weights = Dead) %>%
  mutate(time = Assay_Day, status = 1)

# expand rows for surviving beetles
inj_survivor_rows <- injection %>%
  group_by(CPB_Group, Treatment) %>%
  slice_max(Assay_Day, with_ties = FALSE) %>%
  ungroup() %>%
  filter(Alive > 0) %>%
  uncount(weights = Alive) %>%
  mutate(time = Assay_Day, status = 0)

# combine into survival dataset
inj_survival_long <- bind_rows(inj_death_rows, inj_survivor_rows)

# ensure factor levels are in your desired order
treatment_levels <- c(
  "Control_Uninjected", "Control_H2O", "Control_dsRNA",
  "GPA2_Low", "GPA2_High", "LGR1_Low", "LGR1_High"
)

inj_survival_long <- inj_survival_long %>%
  mutate(Treatment = factor(Treatment, levels = treatment_levels))

# kaplan-meier fit
inj_km_fit <- survfit(Surv(time, status) ~ Treatment, data = inj_survival_long)

# injection palette
injection_palette <- c(
  "Treatment=Control_Uninjected" = "#A8E6A3",
  "Treatment=Control_H2O"        = "#81C784",
  "Treatment=Control_dsRNA"      = "#4CAF50",
  "Treatment=GPA2_Low"           = "#81D4FA",
  "Treatment=GPA2_High"          = "#0288D1",
  "Treatment=LGR1_Low"           = "#CE93D8",
  "Treatment=LGR1_High"          = "#8E24AA"
)

ggsurvplot(
  inj_km_fit,
  data = inj_survival_long,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  xlim = c(0, 28),
  break.time.by = 2,
  palette = injection_palette,
  title = "Kaplan-Meier Survival by Treatment",
  xlab = "Assay Day",
  ylab = "Survival Probability"
)

# test for significance with log-rank test
survdiff(Surv(time, status) ~ Treatment, data = inj_survival_long)

# specific comparisons
pairwise_survdiff(Surv(time, status) ~ Treatment, data = inj_survival_long,
                  p.adjust.method = "BH")
pairwise_survdiff(Surv(time, status) ~ Treatment, data = inj_survival_long,
                  p.adjust.method = "bonferroni")
```

# Dataset for Development Analyses:

```{r convert injection dataset for development modelling}
library(dplyr)
library(tidyr)
library(purrr)

# load and clean raw injection data
transformed_inj <- read.csv("Injection Trial Data.csv") %>%
  select(-`Weight..g.`, -`X..Defoliation`)

# pivot life stage columns to long format
transformed_inj_long <- transformed_inj %>%
  pivot_longer(
    cols = c(Alive, Dead, Incapacitated, Pupating, Adults),  
    names_to = "Life_Stage",  
    values_to = "Count"       
  )

# expand each row so that each beetle is a separate row
transformed_inj_expanded <- transformed_inj_long %>%
  uncount(weights = Count)

# define numeric life stage codes
life_stage_codes <- c(
  Alive = 1,
  Incapacitated = 2,
  Pupating = 3,
  Adults = 4,
  Dead = 5
)

# prepare wide-format daily counts by group
inj_counts_wide <- transformed_inj %>%
  mutate(across(c(Alive, Dead, Incapacitated, Pupating, Adults), as.integer)) %>%
  arrange(`CPB.Group`, `Assay.Day`)

# simulate beetle-level life stage assignment day by day
inj_progression_list <- list()

for (group in unique(inj_counts_wide$CPB.Group)) {
  group_data <- inj_counts_wide %>%
    filter(CPB.Group == group) %>%
    arrange(`Assay.Day`)
  
  beetles <- tibble(Beetle_ID = 1:10, Stage = rep(1, 10))
  group_history <- list()
  
  for (i in seq_len(nrow(group_data))) {
    day <- group_data[i, ]
    target_counts <- as.integer(day[c("Alive", "Incapacitated", "Pupating", "Adults", "Dead")])
    names(target_counts) <- c("Alive", "Incapacitated", "Pupating", "Adults", "Dead")
    
    remaining <- beetles
    assigned <- NULL
    
    for (stage_name in rev(names(life_stage_codes))) {
      code <- life_stage_codes[stage_name]
      n_needed <- target_counts[stage_name]
      eligible <- remaining %>% filter(Stage <= code)
      selected <- if (nrow(eligible) >= n_needed) slice_head(eligible, n = n_needed) else eligible
      
      if (nrow(selected) > 0) {
        selected$Stage <- code
        assigned <- bind_rows(assigned, selected)
        remaining <- anti_join(remaining, selected, by = "Beetle_ID")
      }
    }
    
    if (!is.null(assigned)) {
      day_result <- assigned %>%
        mutate(
          CPB.Group = group,
          Assay.Day = day$Assay.Day,
          Date = day$Date,
          Life_Stage = names(life_stage_codes)[Stage]
        )
      group_history[[i]] <- day_result
      beetles <- assigned
    }
  }
  
  inj_progression_list[[group]] <- bind_rows(group_history)
}

# combine all beetles across all groups and days
inj_beetle_progression <- bind_rows(inj_progression_list) %>%
  arrange(CPB.Group, Beetle_ID, Assay.Day)

# build complete skeleton across all assay days up to 27, default from above code was shorter
max_day <- 27  # or use: max(transformed_inj$Assay.Day, na.rm = TRUE)
inj_skeleton <- transformed_inj %>%
  distinct(CPB.Group) %>%
  crossing(Beetle_ID = 1:10, Assay.Day = 1:max_day)

# join observed progression to skeleton and fill forward stage
inj_beetle_progression_filled <- inj_skeleton %>%
  left_join(inj_beetle_progression, by = c("CPB.Group", "Assay.Day", "Beetle_ID")) %>%
  arrange(CPB.Group, Beetle_ID, Assay.Day) %>%
  group_by(CPB.Group, Beetle_ID) %>%
  fill(Stage, .direction = "down") %>%
  ungroup() %>%
  mutate(
    Life_Stage = names(life_stage_codes)[Stage],
    Life_Stage = factor(Life_Stage, levels = names(life_stage_codes), ordered = TRUE)
  )

# final clean output with globally unique Beetle_ID
inj_beetle_prog <- inj_beetle_progression_filled %>%
  select(CPB.Group, Beetle_ID, Assay.Day, Stage, Life_Stage) %>%
  arrange(CPB.Group, Beetle_ID, Assay.Day)

# assign unique ID across all beetles
unique_beetles <- inj_beetle_prog %>%
  distinct(CPB.Group, Beetle_ID) %>%
  arrange(CPB.Group, Beetle_ID) %>%
  mutate(Global_Beetle_ID = row_number())

inj_beetle_prog <- inj_beetle_prog %>%
  left_join(unique_beetles, by = c("CPB.Group", "Beetle_ID")) %>%
  select(CPB.Group, Global_Beetle_ID, Assay.Day, Stage, Life_Stage) %>%
  rename(Beetle_ID = Global_Beetle_ID)

# preview and export
head(inj_beetle_prog)
write.csv(inj_beetle_prog, "inj_data_transformed.csv", row.names = FALSE)

# add in from, to, start, stop for transition specific modelling
# add a run ID that increments each time the stage changes
inj_transitions_grouped <- inj_beetle_prog %>%
  arrange(Beetle_ID, Assay.Day) %>%
  group_by(Beetle_ID) %>%
  mutate(
    stage_change = Stage != lag(Stage, default = first(Stage)),
    run_id = cumsum(stage_change)
  ) %>%
  group_by(Beetle_ID, run_id) %>%
  summarise(
    CPB.Group = first(CPB.Group),
    from = first(Stage),
    start = min(Assay.Day),
    stop = max(Assay.Day) + 1,  #end of the interval
    .groups = "drop"
  ) %>%
  group_by(Beetle_ID) %>%
  mutate(to = lead(from)) %>%  #get the next state
  ungroup()

# #add Life_Stage labels if desired
# inj_transitions_grouped <- inj_transitions_grouped %>%
#   mutate(
#     from_label = factor(from, levels = 1:5, labels = names(life_stage_codes)),
#     to_label   = factor(to,   levels = 1:5, labels = names(life_stage_codes))
#   )

inj_transitions_grouped <- inj_transitions_grouped %>%
  select(-run_id)

head(inj_transitions_grouped)
write_csv(inj_transitions_grouped, file = "inj_data_start-stop.csv")
```

Data transformation success, yay! 

Disease progression model needs all stages (1-2, 1-3, 1-4, 1-5, etc...) status and incidence indicate whether or not the event happened. Day of progression through one stage is the day another stage can potentially start.

Reminder of lifestage codes:
  Alive = 1,
  Incapacitated = 2,
  Pupating = 3,
  Adults = 4,
  Dead = 5
  
# Development Code:

```{r set-up data}
# read in the data
library(readr)
inj_data_start_stop <- read_csv("Injection Trial Data.csv")
beetle_stages <- inj_data_start_stop

# check out the data
summary(beetle_stages)
str(beetle_stages)

# extract the treatments
library(stringr)
split_cols <- as.data.frame(str_split_fixed(beetle_stages$CPB.Group, "[.]", 2))
colnames(split_cols) <- c("treatment", "rep")

# join them onto the data
beetle_stages_appended <- cbind(beetle_stages, split_cols)

# add in a "time in state" variable, B(t)
beetle_stages_appended$Bt <- beetle_stages_appended$stop-beetle_stages_appended$start
```

```{r examples of certain transitions}
library(dplyr)
# create a dataset to join onto

# study only 1 to 2
beetle_12 <- beetle_stages_appended %>% 
  filter(from == 1) %>%
  select(Beetle_ID, treatment, rep, from, to, start, stop) %>% 
  mutate(to_for_status = 2) %>% 
  mutate(status = ifelse(to_for_status == to, 1, 0)) %>% 
  # Add a line that says 0 if.na. This changed the coutcome
  mutate(status = ifelse(is.na(status), 0, status)) %>%
  select(-to) %>% 
  rename(to = to_for_status)

# study only 2 to 5
beetle_25 <- beetle_stages_appended %>% 
  filter(from == 2) %>%
  select(Beetle_ID, treatment, rep, from, to, start, stop) %>% 
  mutate(to_for_status = 5) %>% 
  mutate(status = ifelse(to_for_status == to, 1, 0)) %>% 
  mutate(status = ifelse(is.na(status), 0, status)) %>%
  select(-to) %>% 
  rename(to = to_for_status)

library("survival")

### 1->2: remission to relapse
obj12 <- coxph(Surv(start,stop,status)~ treatment
               ,data=beetle_12,
            subset=((from==1)&(to==2))
            )
summary(obj12)

### try running 2->5: Incapacitated to death
obj25 <- coxph(Surv(start,stop,status)~ treatment
               ,data=beetle_25,
            subset=((from==2)&(to==5))
            )
summary(obj25)
```

```{r model function}
# Write a function
#' A function that computes the status correctly (was there a transition) and when it occured
#'
#' @param from_stage - start stage
#' @param to_stage - end stage
#' @param df_in - input dataframe
#'
#' @returns - a list with the model and the dataset used to model it 
#' @export
#'
#' @examples - model_1_to_2 <- multi_stage_model(from_stage = 1, to_stage = 2)
multi_stage_model <- function(from_stage, to_stage, df_in = beetle_stages_appended, quad_3_4_flag = 1) {
  # Extract the dataset
  model_df <- {{df_in}} %>% 
    # Only select the from stages that make sense
    filter(from == from_stage) %>%
    # Select only the variables needed in the model
    select(Beetle_ID, treatment, rep, from, to, start, stop, Bt) %>% 
    # Append a variable that's the target
    mutate(to_for_status = to_stage) %>% 
    # Create a status variable if the model actually moved
    mutate(status = ifelse(to_for_status == to, 1, 0)) %>% 
      # Add a line that says 0 if.na
    mutate(status = ifelse(is.na(status), 0, status)) %>%
    # De-select to, as this contains irrelevant "to" values. 
    # The only one that matters is the "to_stage". If that doesn't occur, then the status is 0.
    select(-to) %>% 
    # Rename to
    rename(to = to_for_status)
  # Return the model
  library("survival")
  # Fit one of two models, depending on the "from" stage. If from = 1, don't include current time
  if (from_stage == 1){
    model <- coxph(Surv(start,stop,status)~ treatment, data=model_df,
            subset=((from==from_stage)&(to==to_stage)))
    print("Note: This model doesn't include the time in the current state as it starts from stage 1.")
  } else if ((from_stage == 3) & (to_stage == 4) & (quad_3_4_flag == 1)){
    model <- coxph(Surv(start,stop,status)~ Bt+I(Bt^2)+treatment, data=model_df,
            subset=((from==from_stage)&(to==to_stage)))
    print("Note: This model includes the time in the current state, plus the square of the current time, as it does not start from stage 1.")
  } else{
    model <- coxph(Surv(start,stop,status)~ Bt+treatment, data=model_df,
            subset=((from==from_stage)&(to==to_stage)))
    print("Note: This model includes the time in the current state as it does not start from stage 1.")
  }
  # Return the model and the model_df
  return(list(model = model, model_df = model_df))
}
```

```{r modelling and summaries}
# run the model
model_1_to_2 <- multi_stage_model(from_stage = 1, to_stage = 2)
model_1_to_3 <- multi_stage_model(from_stage = 1, to_stage = 3)
model_1_to_4 <- multi_stage_model(from_stage = 1, to_stage = 4) # Should be nonsense, as this had no events
model_1_to_5 <- multi_stage_model(from_stage = 1, to_stage = 5)

# start from 2
model_2_to_5 <- multi_stage_model(from_stage = 2, to_stage = 5)

# start from 3
model_3_to_2 <- multi_stage_model(from_stage = 3, to_stage = 2) # No events
model_3_to_4 <- multi_stage_model(from_stage = 3, to_stage = 4) # This uses the quadratic time model
model_3_to_4_no_quad <- multi_stage_model(from_stage = 3, to_stage = 4, quad_3_4_flag = 0) # This uses the linear time model
model_3_to_5 <- multi_stage_model(from_stage = 3, to_stage = 5) 

# start from 4
model_4_to_2 <- multi_stage_model(from_stage = 4, to_stage = 2) # No events?
model_4_to_5 <- multi_stage_model(from_stage = 4, to_stage = 5) 

# look at summaries
model_1_to_2$model # Not significant (incapacitated)
model_1_to_3$model # Significant
model_1_to_5$model # Not
model_2_to_5$model # Significant
model_3_to_2$model # No events
model_3_to_4$model # Significant, includes quadratic term
model_3_to_5$model # Some of these p-values are high because there are no events that occured. For instance, no beetles in C2 transitioned to adulthood.
model_4_to_2$model # No events
model_4_to_5$model # Same as 3 to 5.
```

```{r compare model 3-4, fix isssue}
# Compare 3 to 4
model_3_to_4$model 
model_3_to_4_no_quad$model # No issue

BIC(model_3_to_4$model)
BIC(model_3_to_4_no_quad$model)
# The BIC for the no quad model is lower, but it is clearly misspecified per the visualization.
```

```{r visualize raw data as check}
library(ggplot2)

# look at three to four dataset
model_1_to_3$model_df %>% 
  # filter(status == 1) %>% 
  mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = treatment, y = difference, color = as.factor(status))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

bad_df <- model_1_to_3$model_df %>% 
  filter(status == 1) %>%
  mutate(difference = as.numeric(stop - start))

# mod <- lm(difference ~ as.factor(treatment), df = bad_df)
bad_mod <- lm(bad_df$difference ~ bad_df$treatment)
summary(bad_mod)

summary(model_1_to_3$model)

# look at three to four dataset
model_3_to_4$model_df %>% 
  # filter(status == 1) %>% 
  mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = treatment, y = difference, color = as.factor(status))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at three to five dataset
model_3_to_5$model_df %>% 
  # filter(status == 1) %>% 
  mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = treatment, y = difference, color = as.factor(status))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at four to five dataset
model_4_to_5$model_df %>% 
  # filter(status == 1) %>% 
  mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = treatment, y = difference, color = as.factor(status))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at time in stage vs. probability of survival
model_2_to_5$model_df %>% 
  # filter(status == 1) %>% 
  # mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = Bt, y = status, color = as.factor(treatment))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at time in stage vs. probability of survival
model_3_to_4$model_df %>% 
  # filter(status == 1) %>% 
  # mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = Bt, y = status, color = as.factor(treatment))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at time in stage vs. probability of survival
model_3_to_5$model_df %>% 
  # filter(status == 1) %>% 
  # mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = Bt, y = status, color = as.factor(treatment))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))

# look at time in stage vs. probability of survival
model_4_to_5$model_df %>% 
  # filter(status == 1) %>% 
  # mutate(difference = stop - start) %>% 
  #
  ggplot(aes(x = Bt, y = status, color = as.factor(treatment))) +
  geom_point(position = position_jitter(width = 0.2, height = 0))
```

```{r}
library(emmeans)
EMM <- emmeans(model_3_to_4$model, "treatment")   
EMM    

### pairwise comparisons
contrast(EMM, "pairwise")    # or pairs(EMM)
contrast(EMM, "pairwise",  type = "response")   # or pairs(EMM)
```
